---
title: "Tunis day 1 practical"
author: "daniel.lundin@lnu.se"
date: "20 november 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tunis bioinformatics course, day 1 practical

The task for the first day's practical was to plot monthly
means of relative abundances coloured by the main phyla.

This documents *one* way of performing this.

The data consists of two tables:

1. A wide matrix with counts: taxa as rows and samples as
columns

2. A table with sample information including year, date and
season

We're going to use the Tidyverse to do the analysis.

```{r}
library(tidyverse)
```

We start by reading the two tables.

First, read the counts matrix and convert to long format.

```{r read-counts}
counts <- read_tsv(
  'tunis_day1_practical.matrix.tsv.gz',
  col_types = cols(
    .default = col_integer(), 
    seqid = col_character(),
    taxonhier = col_character()
  )
) %>%
  # Make it long, and filter away all zeroes
  gather(sample, count, 3:40) %>%
  filter(count > 0) %>%
  # Calculate relative abundances per sample
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>%
  ungroup() %>%
  # Separate the taxonomy hierarchy string
  separate(
    taxonhier, c(
      'domain', 'phylum', 'class', 'order', 'family', 'genus'
    ),
    sep = ';'
  )
```

Then, read the samples data and join with the counts.

```{r}
samplecounts <- read_tsv(
  'tunis_day1_practical.samples.tsv.gz',
  col_types = cols(
    .default = col_character(),
    date = col_date(format = '%Y-%m-%d')
  )
) %>%
  # Select some of the columns and create a "month" column
  transmute(sample, year, season, date, month = lubridate::month(date)) %>%
  inner_join(counts, by='sample')
```

Since we have so many phyla, find the top 9 (the rest will
be "Other phyla").

```{r}
top9phyla <- counts %>%
  group_by(phylum, sample) %>%
  summarise(count = sum(count)) %>%
  summarise(meancount = mean(count)) %>%
  ungroup() %>%
  top_n(9, meancount)
```

Time to calculate mean abundances of the top 9 phyla plus
"Other phyla" per year and month.

```{r monthly-means}
monthly_means <- samplecounts %>%
  # Join in (left) the top 9 phyla and assing "Other phyla"
  # to the phyla not among the 9.
  left_join(
    top9phyla %>% transmute(phylum, mainph = phylum),
    by='phylum'
  ) %>%
  replace_na(list('mainph' = 'Other phyla')) %>%
  # Calculate sums of phyla per sample
  group_by(mainph, year, month, sample) %>%
  summarise(relab = sum(relab)) %>%
  # Calculate the mean over a month (sample is lost from
  # group_by now, since it was last in the list).
  summarise(meanrelab = mean(relab)) %>%
  # It's always a good idea to ungroup() when you're done
  ungroup()
```

Now we can plot

```{r}
monthly_means %>%
  ggplot(
    aes(x = month, y = meanrelab, fill = mainph)
  ) +
  # Barplot
  geom_col() +
  # A discrete, i.e. integer, x scale from 1 to 12
  scale_x_discrete(
    limits = 1:12
  ) +
  # Nicer (?) colours
  scale_fill_brewer('', palette = 'Set3') +
  xlab('Month') +
  ylab('Mean relative abundance') +
  # Change the default setting for theme to bw
  theme_bw() + 
  theme(
    legend.position = 'bottom' # Colour legend at the bottom
  )
```

