---
title: "Tunis day 1 demo"
author: "daniel.lundin@lnu.se"
date: "20 november 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
```

# Basic handling of 16S data

## Read the table with counts

```{r}
lmocounts = read_tsv(
  '../LMO/16S/dada2/LMO.16S.dada2.tsv.gz',
  col_types = cols(
    seqid = col_character(),
    sample = col_character(),
    count = col_integer()
  )
)
```

A table with sums.

```{r sums}
lmocounts %>% group_by(sample) %>%
  summarise(sumcount = sum(count), mean = mean(count), n = n())
```

Very basic plot of counts per sample.

```{r}
lmocounts %>% group_by(sample) %>%
  summarise(count = sum(count)) %>%
  ggplot(aes(x=sample, y = count)) +
  geom_point()
```

Let's do a boxplot of everything.

```{r}
lmocounts %>%
  ggplot(aes(x = sample, y = count)) +
  geom_boxplot() +
  scale_y_log10()
```

Read the samples information table, so we have the filter fraction and
date for each sample.

```{r read-sample-data}
samples = read_tsv(
  '../LMO/16S/dada2/LMO.samples.tsv',
  col_types = cols(
    .default = col_character(), 
    date = col_date(format="%Y%m%d")
  )
)
```

Now, we can subset the counts table to just prefiltered samples from
2011-13.

```{r subset-samples}
prefiltered <- samples %>% 
  filter(
    filter_fraction == '3-0.2', 
    lubridate::year(date) %in% c(2011, 2012, 2013)
  ) %>%
  select(sample, date) %>%
  inner_join(lmocounts, by = 'sample') %>%
  group_by(sample) %>% mutate(relab = count/sum(count)) %>% ungroup()
```

## Summarise 2011-2013



```{r prefilt-2011-13-boxplot}
prefiltered %>% 
  ggplot(aes(x = sample, y = relab)) +
    geom_boxplot() +
    scale_y_log10() +
    theme(
      axis.text.x = element_text(angle = 60, hjust = 1)
    )
```

Let's read in the taxonomy, allowing us to use that to e.g. colour
plots.

```{r read-taxonomy}
taxonomy = read_tsv(
  '../LMO/16S/dada2/LMO.16S.wang.silva.hier.tsv.gz',
  col_types = cols(.default = col_character()) 
)
```

Identify the top 9 phyla.

```{r top9phyla}
top9phyla <- prefiltered %>% 
  inner_join(taxonomy %>% select(seqid, phylum), by = 'seqid') %>%
  group_by(phylum, sample) %>%
  summarise(count = sum(count)) %>%
  summarise(meancount = mean(count)) %>%
  arrange(desc(meancount)) %>%
  top_n(9, meancount)
```

Plot relative abundances with taxonomy.

```{r relab-taxonomy-dist}
prefiltered %>%
  inner_join(taxonomy %>% select(seqid, phylum), by = 'seqid') %>%
  left_join(
    top9phyla %>% transmute(phylum, major_ph = phylum), by = 'phylum'
  ) %>%
  replace_na(list('major_ph' = 'Other phyla')) %>%
  ggplot(aes(x = as.character(date), y = relab)) +
  geom_boxplot() +
  geom_jitter(aes(colour = major_ph), size = 1) +
  scale_y_log10() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = 'bottom'
  ) +
  facet_wrap(~lubridate::year(date), scales = 'free_x', ncol = 1)
```

Make the data wide.

```{r spread-gather}
wide2long <- prefiltered %>% select(-count) %>% 
  spread(sample, relab, fill = 0) %>%
  gather(sample, relab, LMO.16SDNA.20110325.3.0.2:LMO.16SDNA.20131211.3.0.2) %>%
  filter(relab > 0)
```

Unite and separate.

```{r}
unite.spread = taxonomy %>% 
  unite(thier, domain:genus, sep=';') %>%
  separate(thier, c('domain', 'ph', 'c', 'o', 'f', 'g'), sep=';')
```


