---
title: "Day 3 and 4"
author: "daniel.lundin@lnu.se"
date: "22 november 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
library(vegan)
```

## Read counts and the mapping table

The raw counts table contains the *sequence*, which is long and not
present in the taxonomy table. To get the *seqid*, which is the key
in the taxonomy table, we need to join with the name2seq table.

In the process, we also calculate the relative abundance.

```{r}
counts = read_tsv(
  'dada2.cleaned.merged.bimeras.tsv.gz',
  col_types = cols(.default = col_character(), count = col_integer())) %>%
  inner_join(
    read_tsv(
      'dada2.cleaned.merged.bimeras.name2seq.tsv.gz',
      col_names = c('seqid', 'sequence'),
      col_types = cols(.default = col_character())
    ),
    by = c('seq' = 'sequence')
  ) %>%
  select(-seq) %>%
  group_by(sample) %>%
  mutate(relab = count/sum(count)) %>%
  ungroup()
```

Plot the number of counts per sample (the *sequencing depth*) and the
relative abundances (multiplied by 10000). The latter is just to check
that we calculated the relative abundances correctly.

```{r plot-seq-depth}
counts %>% group_by(sample) %>%
  summarise(count = sum(count), relab = 10000 * sum(relab)) %>%
  tidyr::gather(mtype, meas, count, relab) %>%
  ggplot(aes(x = sample, y = meas, shape = mtype)) +
  geom_point()
```

What does the distribution of populations within samples look like?

```{r}
counts %>%
  ggplot(aes(x = sample, y = relab)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1) +
  scale_y_sqrt()
```

What about the distribution of populations over samples?

```{r}
toppops = counts %>%
  group_by(seqid) %>%
  summarise(mean_relab = mean(relab)) %>%
  ungroup() %>%
  top_n(60, mean_relab)

counts %>%
  inner_join(toppops, by = 'seqid') %>%
  ggplot(aes(x = seqid, y = relab)) +
  geom_boxplot()
```

Let's read and lock at the taxonomy.

```{r}
taxonomy = read_tsv(
  'dada2.cleaned.merged.bimeras.nr.wang.taxonomy.tsv.gz',
  col_names = c('seqid', 'taxhier'),
  col_types = cols(.default = col_character())
) %>%
  mutate(
    taxhier = gsub('\\([0-9]+\\)', '', taxhier)
  ) %>%
  separate(
    taxhier, 
    c( 'domain', 'phylum', 'class', 'order', 'family', 'genus' ),
    sep = ';', extra = 'drop'
  )
```

Plot all taxa without counts.

```{r}
taxonomy %>% 
  ggplot(aes(x = phylum)) +
  geom_bar() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

Show a heatmap of class occurrence in samples.

```{r fig.height=12, fig.caption = "Heatmap of classes and samples."}
counts %>% select(seqid, relab, sample) %>%
  inner_join(taxonomy %>% select(seqid, class), by='seqid') %>%
  group_by(sample, class) %>%
  summarise(relab = sum(relab)) %>%
  ungroup() %>%
  ggplot(aes(x = sample, y = class, fill = relab)) +
  geom_tile() +
  scale_fill_viridis() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

# Show the vegan package

```{r prepare-vegan-data}
cm = counts %>% select(-relab) %>%
  spread(seqid, count, fill = 0) %>%
  as.data.frame() %>%
  tibble::remove_rownames() %>%
  tibble::column_to_rownames('sample') %>%
  as.matrix()
```

```{r}
rc <- rarecurve(cm, step=100)
```

```{r calc-min}
smallest <- counts %>%
  group_by(sample) %>% summarise(c = sum(count)) %>%
  ungroup() %>%
  summarise(minc = min(c)) %>%
  pull(minc)
```

```{r rarefy}
rare.counts <- rrarefy(cm, smallest)
```

```{r}
tibble::tibble(
  sample = rownames(rare.counts),
  specnumber = specnumber(rare.counts) ,
  shannon = diversity(rare.counts)
) %>%
  ggplot(aes(x = sample, y = shannon)) +
  geom_point()
```

```{r}
nmds.2 = metaMDS(rare.counts, k = 2)
nmds.3 = metaMDS(rare.counts, k = 3)
```

```{r}
samples <- counts %>% distinct(sample) %>%
  separate(
    sample,  c('a','b','date','d'), 
    sep='\\.',  extra='merge', remove = FALSE
  ) %>%
  transmute(
    sample,  date = as.Date(date, "%Y%m%d"),
    month = lubridate::month(date),
    season = case_when(
      month %in% c(12, 1, 2) ~ 'winter',
      month %in% 3:5         ~ 'spring',
      month %in% 6:8         ~ 'summer',
      TRUE                   ~ 'autumn'
    ),
    year = lubridate::year(date)
  )
```

Plot the NMDS with ggplot2.

```{r nmds3-1-2}
scores(nmds.3) %>%
  as.data.frame() %>%
  tibble::rownames_to_column('sample') %>%
  inner_join(samples, by = 'sample') %>%
  ggplot(aes(x = NMDS1, y = NMDS2, colour = season, shape = as.character(year))) +
  geom_point() +
  geom_text(aes(label = month))
```

```{r nmds3-1-3}
scores(nmds.3) %>%
  as.data.frame() %>%
  tibble::rownames_to_column('sample') %>%
  inner_join(samples, by = 'sample') %>%
  ggplot(aes(x = NMDS1, y = NMDS3, colour = season, shape = as.character(year))) +
  geom_point() +
  geom_text(aes(label = month))
```
