---
title: "Day 4 phylum plot"
author: "Students"
date: "23 november 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

We will need a couple of tidyverse libraries.

```{r libraries}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lubridate))
```


# The data

We have three tables, one containing counts per population (*sequence*) and
sample, one containing a map from sequence to seqid and one containing taxonomy 
per population.

```{r read-counts}
counts <- read_tsv(
  'dada2.cleaned.merged.bimeras.tsv.gz',
  col_types = cols(.default = col_character(), count = col_integer())
)
```

```{r read-seq2seqid-map}
seq2seqid <- read_tsv(
  'dada2.cleaned.merged.bimeras.name2seq.tsv.gz',
  col_names = c('seqid', 'seq'),
  col_types = cols(.default = col_character())
)
```


```{r read-taxonomy}
taxonomy <- read_tsv(
  'dada2.cleaned.merged.bimeras.nr.wang.taxonomy.tsv.gz',
  col_names = c('seqid', 'thier'),
  col_types = cols(.default = col_character())
) %>%
  mutate(thier = gsub("\\([0-9]+\\)", "", thier)) %>%
  separate(thier, c('Domain', 'Phylum', 'Class', 'x'), sep = ';', extra='merge') %>%
  mutate(nphylum = ifelse(Phylum == 'Proteobacteria', Class, Phylum))
```

## 1. Join the counts with the map and deselect the sequence
```{r}
counts = counts %>%
  inner_join(seq2seqid, by = 'seq') %>%
  select(-seq) %>%
  group_by(sample) %>%
  mutate(relab = count/sum(count)) %>%
  ungroup()
```

## 2. Calculate phylum/class, join with counts and find top 10
```{r}
top10 = taxonomy %>%
  inner_join(counts, by = 'seqid') %>%
  group_by(nphylum, sample) %>%
  summarize(sumrelab = sum(relab)) %>%
  summarise(maxrelab = max(sumrelab), meanrelab = mean(sumrelab)) %>%
  top_n(10, maxrelab) 
```

## 3. Plot barchart with top 10 + other phyla relative abundances per sample
```{r}
taxonomy %>%
  inner_join(counts, by = 'seqid') %>%
  inner_join(top10, by = 'nphylum') %>%
  group_by(sample, nphylum) %>%
  summarise(relab = sum(relab)) %>%
  ungroup() %>%
  ggplot(aes(x=sample, y = relab, fill = nphylum))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```


